The Categorical Structure of Profunctor Optics

### 2017-11-03 paper.tex
Add reference for doctrinal adjunction
Typoes etc

\begin{theorem}
If $(\C, \otimes, I)$ is a symmetric monoidal category, then $\PreOptic_\otimes$ is also symmetric monoidal. \todo{braided?}
\end{theorem}
\begin{proof}
On objects, we define $(S, S') \otimes (T, T') := (S \otimes T, S' \otimes T')$. The unit is given by the object $(I, I)$. The action of $\otimes$ on morphisms is induced by
\begin{align*}
&\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(T, N \otimes B) \times \C(N \otimes B', T')\\
\to \quad&\C(S \otimes T, M \otimes A \otimes N \otimes B) \times \C(M \otimes A' \otimes N \otimes B', S' \otimes T') && \text{(action of $\otimes$ on morphisms in $\C$)}\\
\to \quad&\C(S \otimes T, M \otimes N \otimes A \otimes B) \times \C(M \otimes N \otimes A' \otimes B', S' \otimes T') && \text{($\otimes$ in $\C$ is symmetric)}\\
\to \quad&\int^{M \in \C} \C(S \otimes T, M \otimes A \otimes B) \times \C(M \otimes A' \otimes B', S' \otimes T') && \text{(inclusion into coend)}
\end{align*}
yielding a function $\PreOptic_\otimes((S, S'), (A, A')) \times \PreOptic_\otimes((T, T'), (B, B')) \to \PreOptic_\otimes((S \otimes T, S' \otimes T'), (A \otimes B, A' \otimes B'))$.

@@

Suppose we have two optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$, where $l : R \to MS$ and $l' : S \to NA$, and such that $lr = \phi S$ and $l'r' = \psi A$. Then the composite $\langle (Ml')l, r (Mr')  \rangle$ is also an optic:
\begin{align*}
r (Mr')(Ml')l &= r M(r'l') l = rl = \id_S
\intertext{and}
(Ml')l r (Mr') &= (Ml')(\phi S)(Mr') = (\phi NA) M(l')M(r') = (\phi NA) M(l'r') = (\phi NA) M(\psi A) = \phi(\psi A)
\end{align*}
by the naturality of the action of $\M$.
\end{proof}

@@

Let $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. Verification of the laws is straightforward:
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (t\pi_1 l \times A) \times A) \\
&= r (t\pi_1 l \times A) \pi_{1,3} \\
&= r lr(\pi_1 l \times A) \pi_{1,3} \\
&= r(\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}

\todo{Written out at \cite{IsomorphismLensesPost}, definitely not the first place though.}
\end{proof}

@@

In favourable conditions, we can upgrade this result to show that a lens $S \hto A$ in fact implies that $S \cong M \times A$ for some $M$. This is similar to the result proved in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}

@@

\section{Profunctor Optics}

\subsection{Profunctors}

\todo{recall definitions} We use $I$ to refer to the identity profunctor $\C(-,{=})$, and $\odot$ for horizontal profunctor composition. Profunctor composition is written in diagramatic order.

\subsection{Tambara Modules}
The following section generalises definitions that first appeared in \cite{Doubles}.
\begin{definition}
Suppose $\C$ is acted on by $(\M, \otimes, I)$ and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a natural family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
such that $\alpha_{A,B,I} = \id_{P(A,B)}$, and the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, N \otimes M}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
\end{definition}

@@

\begin{corollary}
$\Pastro_\M$ is oplax monoidal.
\end{corollary}
\begin{proof}
This follows abstractly as a left adjoint of a strong monoidal functor. \todo{reference?}
\end{proof}

@@

Many attempts \cite{ReflectionsOnMonadicLenses} \todo{more} to define effectful lenses have wrapped the result of one or both of $\fget$ and $\fput$ with a monad. This approach is a little ad-hoc; it is not obvious what the laws ought to be and there is no clear generalisation to other varieties of optic. The definition of optic given in this paper suggests we instead set $\C$ to be the Kleisli category for some monad.

\section{Lenses in Linear Types}

>>>>

\begin{theorem}
If $(\C, \otimes, I)$ is a symmetric monoidal category, then $\PreOptic_\otimes$ is also symmetric monoidal. \todo{braided?}
\end{theorem}
\begin{proof}
On objects we define $(S, S') \otimes (T, T') := (S \otimes T, S' \otimes T')$. The unit is given by the object $(I, I)$. The action of $\otimes$ on morphisms is induced by
\begin{align*}
&\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(T, N \otimes B) \times \C(N \otimes B', T')\\
\to \quad&\C(S \otimes T, M \otimes A \otimes N \otimes B) \times \C(M \otimes A' \otimes N \otimes B', S' \otimes T') && \text{(action of $\otimes$ on morphisms in $\C$)}\\
\to \quad&\C(S \otimes T, M \otimes N \otimes A \otimes B) \times \C(M \otimes N \otimes A' \otimes B', S' \otimes T') && \text{($\otimes$ in $\C$ is symmetric)}\\
\to \quad&\int^{M \in \C} \C(S \otimes T, M \otimes A \otimes B) \times \C(M \otimes A' \otimes B', S' \otimes T') && \text{(inclusion into coend)}
\end{align*}
yielding a function $\PreOptic_\otimes((S, S'), (A, A')) \times \PreOptic_\otimes((T, T'), (B, B')) \to \PreOptic_\otimes((S \otimes T, S' \otimes T'), (A \otimes B, A' \otimes B'))$.

@@

Suppose we have two optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$, where $l : R \to MS$ and $l' : S \to NA$, and such that $lr = \phi S$ and $l'r' = \psi A$. Then the composite $\langle (Ml')l, r (Mr')  \rangle$ is also an optic:
\begin{align*}
r (Mr')(Ml')l &= r M(r'l') l = rl = \id_S
\intertext{and}
(Ml')l r (Mr') &= (Ml')(\phi S)(Mr') \\ 
&= (\phi NA) M(l')M(r') \\ 
&= (\phi NA) M(l'r') \\
&= (\phi NA) M(\psi A) \\
&= \phi(\psi A)
\end{align*}
by the naturality of the action of $\M$.
\end{proof}

@@

Let $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. Verification of the laws is straightforward:
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (t\pi_1 l \times A) \times A) \\
&= r (t\pi_1 l \times A) \pi_{1,3} \\
&= r lr(\pi_1 l \times A) \pi_{1,3} \\
&= r(\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}
\end{proof}

This has been written out many times before in the case when $lr$ is the identity \cite{IsomorphismLensesPost} \todo{more}.

@@

In favourable conditions, we can upgrade this result to show that the existence of a lens $S \hto A$ implies that $S \cong M \times A$ for some $M$. This is a more general version of the argument given in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. If $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}

@@

\todo{TODO}

\section{Profunctor Optics}
Let $I = \C(-,{=}) : \C \hto \C$ be the identity profunctor and $\odot$ be profunctor composition, written in diagrammatic order. The following section generalises definitions that first appeared in \cite{Doubles}.
\begin{definition}
Suppose $\C$ is acted on by $(\M, \otimes, I)$ and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a natural family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
such that $\alpha_{A,B,I} = \id_{P(A,B)}$, and the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, N \otimes M}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
\end{definition}

@@

\begin{corollary}
$\Pastro_\M$ is oplax monoidal.
\end{corollary}
\begin{proof}
This follows abstractly as a left adjoint of a strong monoidal functor \cite{Kelly1974}.
\end{proof}

@@

Many attempts \cite{ReflectionsOnMonadicLenses} \todo{more} to define effectful lenses have wrapped the result of one or both of $\fget$ and $\fput$ with a monad. This approach is a little ad-hoc; it is not obvious what the laws ought to be and there is no clear generalisation to other varieties of optic. The definition of optic given here suggests we instead set $\C$ to be the Kleisli category for some monad $T$.

\section{Lenses in Linear Types?}
